\chapter{Introduction to the papers}

\pagebreak
\section{Introduction}

The main scientific part of this thesis consists of five papers. They come in a
sequence to show the research progress within this PhD program. Paper I is
presented in the proceedings of the Computational Methods in Water Resources
(CMWR) conference in Barcelona, 2010. Utilizing the feed-back from that
conference and including more uncertain parameters in the study ended up in
Paper II presented in the proceedings of the ECMOR XII in Oxford, 2010. Paper
III includes a detailed study on the flow responses and effect of relative
permeability on the flow and it is submitted to the International Journal of
Greenhouse Gas Control (IJGGC). Pressure is an important model response during
injection operations. Therefore, a special study is dedicated to pressure
analysis in the system. This is reported in Paper IV, which is submitted to the
IJGGC. Finally, Paper V reports modern stochastic techniques used to perform
detailed quantitative sensitivity analysis and probabilistic risk assessments.
This paper is submitted to special issue of the IJGGC.

\section{Summary of papers}

%
% Paper I
%
\paperitem{I}
{Impact of geological heterogeneity on early-stage $\mbox{CO}_2$ plume
migration}
{%%%------------PAPER I-------------------%%%
\textbf{Summary:}

It is a conventional practice in the context of $\mbox{CO}_2$ storage study to
simplify the geological modeling to achieve an easier force balance
study in the medium. Assuming a homogeneous medium is the first step to quantify
the temporal and spatial scales in $\mbox{CO}_2$ storage problems by a
dimensionless analysis on the analytical solution of flow equation. However, in
practice the real flow performance is very much influenced by geological
heterogeneity. 

We use a set of SAIGUP realizations selected to cover the variability of four
sedimentological and structural parameters. The selected parameters are
lobosity, barriers, aggradation angle, progradation direction. Each
of these parameters varies over three levels, except the progradation direction,
which includes up-dip and down-dip directions. Combining the available 
parameters makes 54 realizations.

$30~\mbox{years}$ of injection and $70~\mbox{years}$ of early migration of
$\mbox{CO}_2$ are simulated and flow responses related to the storage
capacity and leakage risk objectives are defined and calculated from
the simulation results. The responses are reported in scatter plots at the end of
injection and at the end of early migration time. 

This work is specific in examining how heterogeneity influences flow behavior
by using a number of geological realizations. Flow responses defined in
this work are specific to $\mbox{CO}_2$ studies and differ from the responses
used in the original SAIGUP project to study oil recovery. We simulate the
aquifer average pressure,  model boundary fluxes, residual and mobile CO$_2$
saturation, and spatial distribution of connected CO$_2$ volumes. These
responses can be considered to evaluate the site storage capacity and risk of
CO$_2$ leakage to surface. 

\vspace{0.5cm}
\noindent\textbf{Comments:}

This work initially was presented at the  Edinburgh 2010 ACM conference, in a
brief proceedings presentation, and more details of the work are reported in a
proceedings for CMWR 2010 conference in Barcelona. The aim of this paper is to
investigate the impact of geological heterogeneity on a typical $\mbox{CO}_2$
injection problem.

The main concern here is to include realistic geological heterogeneity knowledge
into flow simulation work-flow, which is specific to the $\mbox{CO}_2$ storage
problem. Results of this work conclude that the range of variability in flow
responses indicate the significance of geological heterogeneity in modeling the
$\mbox{CO}_2$ flow. Geological features are ranked by for their impact on each
of the defined flow responses and in particular, aggradation angle has shown a
big impact in most of the responses.

This paper lacks the quantitative sensitivity analysis, which was added in later works. 

\vspace{0.5cm}
\noindent\textbf{Contribution of the candidate:}

Major part of the work is done by the candidate. This includes:
\begin{itemize}
\item Defining the problem.
\item Designing the simulation scenarios.
\item Defining the simulation responses.
\item Performing the simulations  and processing the simulation results.
\item Analyzing the results.
\item Writing the report.
\end{itemize}

The candidate has received helps from the co-authors mainly on analyzing the simulation  results and writing the report. This has been through discussion meetings and review comments on the draft report originally written by the candidate. 
}%%%------------PAPER I-------------------%%%


%
% Paper II
%
\paperitem{II}
{Impact of geological heterogeneity on early-stage $\mbox{CO}_2$ plume
migration: sensitivity study}
{%%%------------PAPER II-------------------%%%
\textbf{Summary:} 

In this paper, we consider faults in addition to the four geological parameters
used in Paper I. In the SAIGUP study, fault modeling is performed in an
intensive variability over structural parameters and transmissibility across the
fault. To keep the work less succinct and conclusive in studying the
dynamics of flow, we fix the structural variability in its medium level and vary
the transmissibility over three main levels: unfaulted, open faults and closed
faults.

The flow responses analyzed are the same as
  defined in Paper I. In addition, we perform a linear sensitivity analysis for
the flow responses
with respect to the geological variations. The outcome of the sensitivity analysis
shows that the flow behavior is highly sensitive to aggradation angle. Barriers
and faulting will also influence the flow responses significantly. In this
work, similar to Paper I, we clearly see the range variation in
flow responses which demonstrates how important it is to model the geological
features accurately. 

\vspace{0.5cm}
\noindent\textbf{Comments:} 

\begin{itemize}

\item \textit{The SAIGUP realizations}
 
% We use the SAIGUP realizations, which are an invaluable set of geological models
% created carefully to represent the heterogeneity variations in shallow-marine
% depositional systems. They are designed for recovery investigations in oil
% fields. The time and spatial scales are smaller in oil and gas field studies
% compared to geological $\mbox{CO}_2$ storage. Gravity plays a major role during
% migration and since it is acting in the vertical direction, special
% considerations must be taken withing flow modeling process.
% 
% The focus in our studies is on injection and early migration phases of
% $\mbox{CO}_2$ storage. Therefore, the size requirements is reduced down to
% capture the region influenced by injection well and the medium that allows for
% evolution of $\mbox{CO}_2$ plumes after injection and before starting the long
% migrations in the geological beddings. However, the pressure disturbances from
% injection well go in size beyond the scales of $\mbox{CO}_2$ distributions.
% Therefore, hydrostatic open boundaries are considered for the SAIGUP
% realizations to minimize the limitations due to model size.

This paper is presented in ECMOR conference in Oxford, 2010. This is a
complementary to proceedings works presented in ACM Edinburgh, 2009 and CMWR in
Barcelona, 2010. 

Topography is a major player in the gravity dominated flow behavior. The SAIGUP
realizations include variability in topography of the geological layering via
structural changes due to faults and also barriers in the model. These are good
enough for early migration when the $\mbox{CO}_2$ and water segregate and plumes
accumulate below cap-rock and start the longer migration. In the long-term
migration, top surface geometry is an important geological parameter and
larger models than the SAIGUP models with a better resolution of the top surface
are needed to get good predictions of the long-term migration phase. This
was considered in the next generation of geological studies performed following
this study \cite{syversveenstudy,nilsen2012impact} under the IGEMS research
project.

\item \textit{Physical assumptions}
 
The work concentrates on how geological heterogeneity impacts the flow
performance. We need to measure the volumetric sweep efficiency of $\mbox{CO}_2$
plumes to evaluate the residual trapping. The most important parameter
is the rock transmissibility, rather than fluid properties. Including more
physics in the modeling will add the computational costs specially when the flow
modeling is used in a sensitivity analysis or risk assessment process.
Therefore, we used simple fluid models for PVT. 

To accelerate the flow, we used linear relative permeability curves. This
increases the flow mobility in the low saturation values. However, in \coo-water system the relative permeability curves are closer to a quadratic function. This is investigated in the next paper (Paper III). Hysteresis effects are
modeled by changing the endpoints in the relative permeability curves and no
scanning-curves is considered here. During injection, the main process is
drainage. After injection, the imbibition process starts and mostly is a
replacement of $\mbox{CO}_2$ by water due to gravity segregation. This justifies
the usage of simple hysteresis model, and more detailed study can be done to
investigate this influence in a quantitative manner. 

\item \textit{Uncertainty considerations}

Our main motivation for using the SAIGUP data was the extensive work that was put into building realistic geological realizations. The geological parameters are changed in value between low and high levels. These values are assumed with the same probability, which is a reasonable start point for sensitivity analysis. In general, this probability might not be uniform, depending to the regional geology of the storage site. 

\end{itemize}
\vspace{0.5cm}
\noindent\textbf{Contribution of the candidate:}

Major part of the work is done by the candidate. This includes:
\begin{itemize}
\item Defining the problem.
\item Designing the simulation scenarios.
\item Defining the simulation responses.
\item Performing the simulations  and processing the simulation results.
\item Analyzing the results.
\item Writing the report.
\end{itemize}

The candidate has received helps from the co-authors mainly on analyzing the simulation  results and writing the report. This has been through discussion meetings and review comments on the draft report originally written by the candidate. 
}%%%------------PAPER II-------------------%%%


%
% Paper III
%
\paperitem{III}
{Impact of geological heterogeneity on early-stage $\mbox{CO}_2$ plume
migration: sensitivity study}
{%%%------------PAPER III-------------------%%%
\textbf{Summary:}

In this paper, we use the same setup that is used in Paper II. Five
geological features are examined, which are lobosity, barriers, aggradation
angle, progradation direction, and faults. Also, the same injection scenario is
used: $30$ years of $\mbox{CO}_2$ injection and $70$ years of early migration.
The injector is controlled by a constant rate and no pressure constraint is set
to allow for all ranges of pressure, including those that are unrealistic.
Moreover, we define an additional model output that is related to the risk of
CO$_2$ leakage through any breakings in the cap-rock.

We examine the influence of simplifying assumptions considered in our works
regarding linearity of relative permeability function. We perform a detailed
flow analysis on various geological realizations using two different relative
permeability relations: linear and quadratic functions. The non-linearity in
relative permeability hinders the flow in the lower saturations values of the
displaced phase, i.e., water phase. We discuss the influence of curvature on
$\mbox{CO}_2$ flow dynamics within the aquifer. Finally, we perform a quantitative sensitivity analysis by using the flow simulation results. The sensitivity analysis results suggest that aggradation angle, fault criteria, and progradation direction are the most influential geological parameter in our study.

\vspace{0.5cm}
\noindent\textbf{Comments:}

Results show that using linear relative permeability function ends up in
conservative conclusions with respect to $\mbox{CO}_2$ distributions in the
domain in terms of storage safety. Since computational costs are much lower for
linear relative permeability scenario than quadratic relative permeability, it
sounds a good strategy to perform sensitivity analysis by using linear relative
permeability function on flow responses that are based on $\mbox{CO}_2$
distribution in the domain.

However, dramatic pressure build-up can happen in the medium during injection
in the quadratic relative permeability scenario. This suggests that for
pressure studies we must use more realistic relative permeability functions.
 Following this work, a detailed pressure study with more realistic relative
permeability curves is performed, which is reported in the next paper.

Within one geological realization, injection location can dramatically impact the injectivity of the well. In fact, this is an uncertain parameter in the \coo\ storage operations. Choosing to inject in the river channels or in the permeable homogeneous parts near the shore will enhance the injectivity and the \coo\ sweep efficiency in the medium. On the other hand, injecting in the locations with low permeabilities and pore-volumes can significantly increase the injection pressure, while limiting the transport of \coo\ in the medium. Studying the impact of injection location can be performed by injecting in many different points in one realization and comparing the corresponding flow responses. However, this will considerably increase the number of detailed simulations in the study. 

For the allowed time, we limited our study to a fixed point by injecting via one well in the flank part of the SAIGUP models. This location is selected after qualitative analysis of a detailed study on a homogeneous case. There, we aimed to fulfill the criterion of maximizing the \coo\ storage volume via increasing the vertical travel path towards the structural trap location under the cap-rock. One mitigating strategy for minimizing the effect of injection location can be to inject via several wells in different location in the medium.  

Similar argument applies to the leakage risk study reported here. We use a leakage probability over the cap-rock that can dramatically influence the calculated leakage risk. We take this assumption to simplify the way we introduce the method. 

\vspace{0.5cm}
\noindent\textbf{Contribution of the candidate:}

Major part of the work is done by the candidate. This includes:
\begin{itemize}
\item Defining the problem.
\item Designing the simulation scenarios.
\item Defining the simulation responses.
\item Performing the simulations  and processing the simulation results.
\item Analyzing the results.
\item Writing the report.
\end{itemize}

The candidate has received helps from the co-authors mainly on analyzing the simulation  results and writing the report. This has been through discussion meetings and review comments on the draft report originally written by the candidate.  

}%%%------------PAPER III-------------------%%%


%
% Paper III
%
\paperitem{IV}
{Geological storage of CO$_2$: heterogeneity impact on pressure behavior}
{\textbf{Summary:}

After observing the influence of relative permeability curvature on pressure
response for $\mbox{CO}_2$ injection studies, in this paper we perform a
detailed pressure study on the chosen geological realizations.

Pressure build-up is an important criterion that can determine the success and failure of
$\mbox{CO}_2$ storage operations. Over-pressurized injections can induce
new fractures and open the existing faults and fractures that increases the risk
of leakage for the mobile $\mbox{CO}_2$ in the domain. On the other
hand, the pressure disturbance imposed on the system travels within the
domain beyond the scales of $\mbox{CO}_2$ distribution. If the CO$_2$ is
injected into a saline aquifer connected to fresh water aquifers, the pressure
pulse may result in fresh water contaminations by the brine far from the
injection point. We define specific pressure responses to examine the pressure
disturbance in the system during injection.

Two injection scenarios are examined for the same $160$ geological realizations
setup. In the first scenario, the injector is set to a fixed volumetric rate to
inject the $\mbox{CO}_2$ volume in $30$ years into the domain, allowing for an
unlimited pressure build-up. In the second scenario, a pressure constraint is
set on the injector that results in various rate of injection in different
geological realizations to inject the same amount of $\mbox{CO}_2$ volume
considered in the first injection scenario. 

Pressure response sensitivity study with respect to different geological
features indicates the significance of aggradation angle, progradation
direction, and faults during injection. A probabilistic pressure analysis is
also performed based on the $160$ simulations on the available realizations.

\vspace{0.5cm}
\noindent\textbf{Contribution of the candidate:}

Almost all parts of the work is done by the candidate. This includes:
\begin{itemize}
\item Defining the problem.
\item Designing the simulation scenarios.
\item Defining the simulation responses.
\item Performing the simulations  and processing the simulation results.
\item Analyzing the results.
\item Writing the report.
\end{itemize}

The candidate has received helps from the first adviser mainly on writing the report. This has been through review comments on the draft report originally written by the candidate. 

}


%
% Paper IV
%
\paperitem{V}
{Geological storage of $\mbox{CO}_2$: global sensitivity analysis and risk
assessment using the arbitrary polynomial chaos expansion}
{\textbf{Summary:}

In this paper, we perform a stochastic sensitivity and risk analysis. We obtain
a high resolution global sensitivity and probabilistic study on the flow
responses that are defined and discussed in the previous papers. We choose
barriers, aggradation angle, and
faults from the SAIGUP geological parameters. Faults are considered by changing
the transmissibility value across them, which
is a continuous parameter. One more parameter is added to the study which is
common in the literature and models the external pressure support from other
aquifers attached to the model (regional groundwater effect).

Flow simulation on high resolution variational geology demands a huge
computational costs. To enhance the calculation speed, we use a data-driven
method that does not need to solve the full physical flow equations. We
approximate the flow solver by a response surface method that is a polynomial
and relates the system output to the input with a minimal computational cost.
We use the arbitrary polynomial chaos expansion (aPC) to approximate the
flow responses. The aPC method considers the uncertainty in the input variables.

A global sensitivity analysis is performed by employing Sobol indices that are
based on variances of responses. The method is shown to be robust in problems
of high levels of complexity and non-linearity.  

And finally, we perform a Monte-Carlo process using the approximating polynomial
on a high resolution input variations. This makes it possible to perform a high
resolution probabilistic study on the flow responses. This way, extreme cases
can be identified by probability of occurrence.

\vspace{0.5cm}
\noindent\textbf{Comments:}

To implement our stochastic technique, we choose geological parameters
in this study that can be interpolated between two levels of their values. For
example, it makes sense to use barriers coverage level of $25\%$ between the low
($10\%$) and medium ($50\%$) levels used in the previous studies. Some of the
geological parameters are discrete and can not be interpolated between two
values. For instance, lobosity can only be varied over three points and we can
not define a $1.5$-lobe. 

Having a large number of points in the input values interval requires
intensive geological modelings to be used in the flow simulations. Using the
data-driven polynomial, the approach only needs evaluating the polynomial in the
defined values, and there is no need for full geological modeling except in
the collocation points, i.e., point values that the polynomial coefficients must
be calculated. 

The work reported here is to demonstrate the work-flow of using the aPC for
global sensitivity analysis and probabilistic risk assessment. A normal
work-flow starts by defining the uncertainties in the input parameters and
follows by building the geological models for the aPC collocation points that are based on those uncertainties.
To perform this study on the SAIGUP models that are consistent
with a uniform uncertainty in the geological parameters, with no loss of
generality, we used uniform uncertainty distributions for our study. However,
the aPC method is not limited to uniform uncertainty descriptions.

Geological features are ranked based on the sensitivity analysis results. The
results are in agreement with dynamics of the flow in the aquifer. Aggradation
angle is the most influential parameter, while the regional groundwater has the
least influence in the model responses. The study is not limited to the assumed
uncertainty of input parameters and the conclusion may change for a very
different uncertainty description. 

\vspace{0.5cm}
\noindent\textbf{Contribution of the candidate:}

Major part of the work is done by the candidate. This includes:

\begin{itemize}
\item Defining the problem.
\item Designing the simulation scenarios.
\item Designing the work-flow.
\item Integrating the aPC MATLAB code into the work-flow.
\item Performing the runs and processing the results.
\item Performing the global sensitivity analysis.
\item Performing the risk assessment.
\item Analyzing the results and preparing plots.
\item Writing the report.
\end{itemize}

The candidate received help from the co-authors and others involved in defining the problem and writing the report. The report has gone through extensive reviews, after it was originally written by the candidate. The core of the report, including the formulation of the method, the description of the modeling, the presentation of the results, and discussions on the results are originally written by the candidate. The review comments resulted in additional elaborations on the pressure issue and some discussions on the aPC technique when it is compared to other techniques in the literature. The candidate received support from the co-authors on the aPC directly related comments. 
}
